The evolution of gene expression in the term placenta of viviparous mammals

### 2015-08-28 manuscript/manuscript.Rnw
add more missing references to the manuscript

The quality of reads was evaluated using FASTQC v. 0.9.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}). Following
this, reads were trimmed based on quality in Galaxy using the FASTQ
Trimmer v. 1.0.0 (Blankenberg et al., 2010). All transcriptomes were
then aligned to reference genomes and assembled using the Tuxedo suite
of next-generation sequencing analysis tools. Annotated reference
genomes were downloaded from Ensembl v. 64 (Flicek et al., 2012),
including Homo sapiens GRCh37, Pan troglodytes CHIMP2.1 (6X coverage),
Canis familiaris BROADD2 (7.6X), Bos taurus UMD3.1 (9.5X), Loxodonta
africana loxAfr3 (7X), and Monodelphis domestica BROADO5 (6.8X). The
genome of the closely-related Pan troglodytes was used to map the
transcriptome of Pan paniscus; the separation between the two
recognized Pan species is generally perceived to be ~1 Mya (Prufer et
al., 2012). Reads were aligned to reference genomes using Bowtie v.
0.12.7 (Langmead et al., 2009) and TopHat v. 1.3.3 (Trapnell et al.,
2009). Cufflinks v. 1.2.1 was then used to assemble transcripts and
quantify expression values using fragments per kilobase of exon per
million mapped reads (FPKM) (Trapnell et al., 2010). RNA-Seq metrics
were computed using Picard (http://picard.sourceforge.net/). For
comparison, we also downloaded the placenta Illumina RNA-Seq
transcriptome of a mouse from (Wang et al., 2011). Raw reads were
mapped as above using the Ensembl v. 64 annotation of the Mus musculus
(house mouse) genome (NCBI m37).

@@

To determine enrichment levels of functional categories of genes in
highly-expressed transcripts in each transcriptome, we performed
functional annotation cluster analyses per transcriptome using DAVID
(Huang et al., 2009; Sherman et al., 2007). DAVID ranks functional
annotation clusters according to enrichment score (ES); functional
annotation clusters are considered significantly enriched if ES>1.3,
as this corresponds to p<0.05 (Huang et al., 2009). We executed
analyses with two gene lists per transcriptome: 1) annotated genes
with expression levels of the top 5\% of the frequency distribution of
FPKM and 2) the top 100 genes with the highest FPKM. Due to the
paucity of annotated gene ontology terms for some species, gene lists
were compared using gene ontology terms derived from the human genome.

@@

All 1:1 orthologues present among human, bonobo, mouse, dog, cow,
elephant, and opossum were determined using Ensembl v.64 gene trees
via the Biomart online data management tool (Kasprzyk, 2011). A total
of 6882 1:1 orthologous genes were expressed in at least one species.
For further analysis, we eliminated all genes that did not have at
least one species with an FPKM>10. A list of these 4259 1:1
orthologous genes was retained for future analyses of gene expression
evolution described below.

We used the software OUCH (Ornstein–Uhlenbeck for Comparative
Hypotheses) in the R statistical environment to investigate
significant shifts in gene expression levels over time (Butler and
King, 2004). OUCH institutes an Ornstein-Uhlenbeck (OU) model of
evolution, which allows for different optima for quantitative trait
values along user-specified branches of a phylogenetic tree. The OU
model explicitly accounts for change in quantitative trait values by
introducing two parameters into the model of stochastic Brownian
motion of change in trait value along a phylogenetic tree through
time. These parameters include, ?, or an optimum trait value, and ,
or the strength of selection (interpreted as “pull” toward the
optimum). In addition, the OU model can also integrate evolutionary
change by assuming that each user-specified set of braches has its own
separate OU parameters (see Butler and King, 2004).

Various OU branch models were tested against a model in which no
variation in optima among branches was allowed. The following branches
were specified in separate models to have a different optimum than all
other branches: human, bonobo, mouse, dog, cow, elephant, opossum,
human+bonobo, Euarchontoglires, and Boreoeutheria. Likelihood values
were computed for each branch model and compared to an OU model with
one optimum using likelihood ratio tests with df=1. We filtered out
genes with universally low expression levels, and included only genes
in which expression in at least one species had an FPKM≥10 as
described above. We corrected for multiple testing using the false
discovery rate (FDR) and calculated adjusted p values (q-values) using
the Benjamini-Hochstein correction. We used the time-calibrated
phylogenetic tree of (Meredith et al., 2011) with all species not used
in this analyses pruned.

To determine enrichment of functional categories in lists of genes
that display a significant shift in expression on a specific branch,
we performed functional annotation cluster analyses on each branch
using DAVID (Huang et al., 2009; Sherman et al., 2007). DAVID ranks
functional annotation clusters according to enrichment score (ES);
functional annotation clusters are considered significantly enriched
if ES>1.3, as this corresponds to p<0.05 (Huang et al., 2009). As
above, due to the paucity of annotated gene ontology terms for some
species, gene lists were compared using gene ontology terms derived
from the human genome.

>>>>

The quality of reads was evaluated using FASTQC v. 0.9.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Following this, reads were trimmed based on quality in Galaxy using
the FASTQ Trimmer v. 1.0.0
\cite{Blankenberg.ea2010:ManipulationofFASTQdata}. All transcriptomes
were then aligned to reference genomes and assembled using the Tuxedo
suite of next-generation sequencing analysis tools. Annotated
reference genomes were downloaded from Ensembl v. 64
\cite{Flicek.ea2012:Ensembl2012}, including \textit{Homo sapiens}
GRCh37, \textit{Pan troglodytes} CHIMP2.1 (6X coverage), \textit{Canis
  familiaris} BROADD2 (7.6X), \textit{Bos taurus} UMD3.1 (9.5X),
\textit{Loxodonta africana} loxAfr3 (7X), and \textit{Monodelphis
  domestica} BROADO5 (6.8X). The genome of the closely-related Pan
troglodytes was used to map the transcriptome of Pan paniscus; the
separation between the two recognized Pan species is generally
perceived to be ~1 Mya (Prufer et al., 2012). Reads were aligned to
reference genomes using Bowtie v. 0.12.7\cite{Langmead2009} and TopHat
v. 1.3.3 \cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
Cufflinks v. 1.2.1 was then used to assemble transcripts and quantify
expression values using fragments per kilobase of exon per million
mapped reads (FPKM)
\cite{Trapnell.ea2010:Transcriptassemblyandquantification}. RNA-Seq
metrics were computed using Picard
(\url{http://picard.sourceforge.net/}). For comparison, we also
downloaded the placenta Illumina RNA-Seq transcriptome of a mouse from
\cite{Wang.ea2011:surveyfornovelimprinted}. Raw reads were mapped as
above using the Ensembl v. 64 annotation of the \textit{Mus musculus} (house
mouse) genome (NCBI m37).

@@

To determine enrichment levels of functional categories of genes in
highly-expressed transcripts in each transcriptome, we performed
functional annotation cluster analyses per transcriptome using DAVID
\cite{Huang.ea2009:Systematicandintegrativeanalysis,Sherman.ea2007:DAVIDKnowledgebasegene-centereddatabase}.
DAVID ranks functional annotation clusters according to enrichment
score (ES); functional annotation clusters are considered
significantly enriched if ES>1.3, as this corresponds to p<0.05
\cite{Huang.ea2009:Systematicandintegrativeanalysis}. We executed
analyses with two gene lists per transcriptome: 1) annotated genes
with expression levels of the top 5\% of the frequency distribution of
FPKM and 2) the top 100 genes with the highest FPKM. Due to the
paucity of annotated gene ontology terms for some species, gene lists
were compared using gene ontology terms derived from the human genome.

@@

All 1:1 orthologues present among human, bonobo, mouse, dog, cow,
elephant, and opossum were determined using Ensembl v.64 gene trees
via the Biomart online data management
tool\cite{Kasprzyk2011:BioMartdrivingparadigmchange}. A total of 6882
1:1 orthologous genes were expressed in at least one species. For
further analysis, we eliminated all genes that did not have at least
one species with an $\mathrm{FPKM}\ge10$. A list of these 4259 1:1 orthologous
genes was retained for future analyses of gene expression evolution
described below.

We used the software OUCH (Ornstein–Uhlenbeck for Comparative
Hypotheses) in the R statistical environment to investigate
significant shifts in gene expression levels over
time\cite{Butler.2004:phlogeneticcomparativeanalysis}. OUCH institutes
an Ornstein-Uhlenbeck (OU) model of evolution, which allows for
different optima for quantitative trait values along user-specified
branches of a phylogenetic tree. The OU model explicitly accounts for
change in quantitative trait values by introducing two parameters into
the model of stochastic Brownian motion of change in trait value along
a phylogenetic tree through time. These parameters include, $\theta$,
or an optimum trait value, and $\alpha$, or the strength of selection
(interpreted as “pull” toward the optimum). In addition, the OU model
can also integrate evolutionary change by assuming that each
user-specified set of braches has its own separate OU parameters
\cite{Butler.2004:phlogeneticcomparativeanalysis}.

Various OU branch models were tested against a model in which no
variation in optima among branches was allowed. The following branches
were specified in separate models to have a different optimum than all
other branches: human, bonobo, mouse, dog, cow, elephant, opossum,
human+bonobo, Euarchontoglires, and Boreoeutheria. Likelihood values
were computed for each branch model and compared to an OU model with
one optimum using likelihood ratio tests with df=1. We filtered out
genes with universally low expression levels, and included only genes
in which expression in at least one species had an
$\mathrm{FPKM}\ge10$ as described above. We corrected for multiple
testing using the false discovery rate (FDR) and calculated adjusted
$p$ values (q-values) using the Benjamini-Hochberg
correction\cite{Benjamini1995:bh_procedure}. We used the
time-calibrated phylogenetic tree of
\cite{Meredith.ea2011:ImpactsofCretaceousTerrestrial} with all species
not used in this analyses pruned.

To determine enrichment of functional categories in lists of genes
that display a significant shift in expression on a specific branch,
we performed functional annotation cluster analyses on each branch
using DAVID
\cite{Huang.ea2009:Systematicandintegrativeanalysis,Sherman.ea2007:DAVIDKnowledgebasegene-centereddatabase}.
DAVID ranks functional annotation clusters according to enrichment
score (ES); functional annotation clusters are considered
significantly enriched if $\mathrm{ES}\ge1.3$, as this corresponds to
$p\le0.05$ \cite{Huang.ea2009:Systematicandintegrativeanalysis}. As
above, due to the paucity of annotated gene ontology terms for some
species, gene lists were compared using gene ontology terms derived
from the human genome.
